Probabilistic forecasting of hourly Emergency Department arrivals

### 2022-12-28 paper/paper.Rmd
comments adressed by bahman

---
title:  Probabilistic forecasting of hourly Emergency Department arrivals
author:
  - name: Bahmna Rostami Tabar
    email: rostami-tabarb@cardiff.ac.uk
    affiliation: Cardiff Business School
    footnote: 12
  - name: Jethro Browell
    email: jethro.browell@glasgow.ac.uk
    affiliation: University of Glasgow
    footnote: 1
  - name: Ivan Svetunkov
    email: i.svetunkov@lancaster.ac.uk
    affiliation: Lancaster University
    footnote: 1
address:
  - code: Cardiff Business School
    address: Cardiff Business School, Cardiff University, UK
  - code: University of Glasgow
    address: School of Mathematics and Statistics, University of Glasgow, UK
  - code: Lancaster University
    address: Centre for Marketing Analytics and Forecasting, Lancaster University, UK
footnote:
  - code: 12
    text: "Corresponding Author"
  - code: 1
    text: "Equal contribution"

abstract: |
   An accurate forecast of Emergency Department (ED) arrivals by an hour of the day is critical to meet patients’ demand. It enables planners to match ED staff to the number of arrivals, redeploy staff, and reconfigure units. This can have many advantages for healthcare staff and the quality of care delivered to patients. In this study, we develop an innovative model based on Generalised Additive Models and an advanced dynamic model based on exponential smoothing to generate an hourly probabilistic forecast of ED arrivals for a prediction window of 48 hours. We compare the forecast accuracy of these models against appropriate benchmarks, including TBATS, Poisson Regression, Prophet, and simple empirical distribution. We use Root Mean Squared Error (RMSE) to examine the point forecast accuracy and assess the forecast distribution accuracy using Quantile Bias, PinBall Score and Pinball Skill Score. Our results indicate that the proposed models outperform their benchmarks for point and probabilistic forecasts. Our developed models can also be generalised to forecast hourly arrivals in other services such as hospitals, ambulances, or clinical desk services.

keywords:  "Emergency Department, Arrivals, Poisson Regression, Probabilistic Forecasting, Generalised Additive Models, Intermittent Exponential Smoothing"

journal: "Annals of Emergency Medicine"
date: "`r Sys.Date()`"
geometry: "top=25mm, left=30mm, right=30mm, bottom=25mm,headsep=10mm, footskip=12mm"
#linenumbers: true
numbersections: true
#output: rticles::elsevier_article
header-includes:  
 \usepackage{adjustbox, float,lscape, bm,amsmath} #use the 'float' package
output:
  bookdown::pdf_book: 
    base_format: rticles::elsevier_article
    citation_package: natbib
bibliography: mybibfile.bib
#biblio-style: authoryear
#csl: elsevier-without-titles.csl
#biblio-style: abbrevnat
---

@@

There are few studies that look at forecasting hourly arrivals in ED and other hospital services using historical time series data and/or predictors such as patient characteristics, weather, holidays and public events. These studies use multiple approaches, including Exponential Smoothing [@SvetunkovAdam], Autoregressive Integrated Moving Average (ARIMA) [@hyndman2021forecasting], Autoregressive Conditional Heteroskedasticity (ARCH) [@bollerslev1994arch], Vector Autoregressive model [@lutkepohl2013vector],  TBATS [@de2011forecasting] and Artificial Neural Networks [@hyndman2021forecasting]. However, most of these studies are limited to only predicting future arrivals as a point forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals. There are few studies that report uncertainty by presenting prediction intervals, but there is no study generating and evaluating the entire forecast distribution of arrivals. Reporting the uncertainty via the forecast distribution is potentially valuable in this setting because the consequences of inadequate staffing are asymmetric. This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient. Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently.

Furthermore, if the impact of under- and overstaffing can be quantified, probabilistic forecasts allow `optimal' decisions that balance the cost associated with under- and overstaffing to be calculated. Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task. Another drawback of existing studies is that the datasets used are relatively small (e.g. time period of 1-2 years), making it challenging to report the forecast accuracy using robust approaches such as time series cross-validation. Such results might not be generalisable. Additionally, most of the forecasting methods used in these publications do not consider the full extent of the multiple seasonality of hourly ED arrivals. Moreover, all previous publications referenced in this paper are not fully reproducible as underlying data and code are not available.

@@

# Research background: hourly ED forecasting {#lit}

Some studies employ models to forecast admissions and arrivals in Emergency Department [@gul2020exhaustive]. The time granularity considered by these studies spans from hourly to yearly. However, given the focus of the paper, we only discuss hourly ED forecasting.

@@

Table \@ref(tab:summarylit) summarize the relevant papers.

@@

According to the studies mentioned above, it can be said that they have shown complications in forecasting hourly patient accident and emergency department visits, and the application of forecasting hourly patient visits is not well established. Some of the studies claimed that the accuracy of forecasting models on hourly accident and emergency department data is low [@boyle2012predicting; @hertzum2017forecasting], while others mentioned that the accuracy of ED hourly forecast is at the acceptable level [@choudhury2020forecasting; @mccarthy2008challenge;  @schweigler2009forecasting].

@@

Data used in this study comprises counts of patients’ arrival times at one of the largest ED units in the UK between April 2014 and February 2019, extracted from the ED administrative database of the hospital. We aggregated the patients’ arrival times to obtain hourly arrivals, which are used in this study. Figure \@ref(fig:hourly-plot-ridge) illustrates the distribution of arrivals for each hour of the day and the day of the week. Although the data is noisy, it reveals some systematic patterns.

@@

```{r seasonplot-dofw, fig.cap= "Seasonal plot: day of week arrivals", fig.align='center',out.width="70%"}
```

Figure \@ref(fig:seasonplot-dofw) illustrates the daily seasonal plot, with the x-axis representing the day-of-week and y-axis the ED arrivals. There is a line for each week (i.e. seven days) from Monday to Saturday, and lines are colour-coded by the week of the year to show the weekly cycle. It is clear that there is a significant jump in patient arrivals on Monday followed by Saturday each week. This indicates that there are significantly more arrivals on Mondays and Saturdays compared to the rest of the week. This might be due to the closure of General Practitioners outpatient clinics over the weekend.

@@

## Naive/Climatology {#climatology}

We start with one of the simplest forecasting approaches used in practice - assuming that in the next few hours, everything will be the same as in the similar hours of a similar day in the past. This is called "Naïve". In our case, given that we need a distribution of values, we will use a modified approach, used in climatology, where the empirical distribution of the hourly arrival time series is used to forecast the future arrival distribution [@la2021new]. We consider the empirical distribution of all available historic data (Benchmark-1) and the empirical distribution of the most recent year of historic data on a rolling basis (Benchmark-2) to capture potential changes in behaviour over time.

@@

The variables (1)--(3) allow modelling the seasonal patterns on the appropriate level of detail throughout the year, while (4) covers the changes in admittance due to calendar events. Finally, (5) is needed to capture the potential phenomenon of change in admittance after the holiday (e.g. people might try not to go to the hospital on Christmas eve and thus go the next day). This model assumes that all these effects are deterministic and do not change over time. Still, the exponentiation in \@ref(eq:PoissonRegression) introduces an interaction effect between dummy variables so that the 3 pm on Monday in January will be different from 3 pm on Monday in July, although the parameters for the hour of day and day of the week are fixed and do not change over time. We use the `alm()` function from the `greybox` package [@Svetunkov2021Greybox] for R [@RTeam2021] for the experiments and denote this model as "Poisson Regression".

@@

Prophet is a forecasting procedure created by Facebook [@taylor2018forecasting] that accounts for multiple seasonality, piecewise trend and holiday effects. Prophet is robust to missing data and shifts in the trend and typically handles outliers well. Prophet works well on daily data seen in Facebook. It is robust and automated, making it easy to learn for beginners. The implementation may be less flexible than other methods. The model itself relies on the Multiple Source of Error state space model, initially proposed by @Kalman1960. The model is incorporated using the corresponding implementation of the Fable package in R. We use the `prophet()` function from the `fable` package [@fable2020]. Note that the input data is assigned with an hourly and daily seasonality.

@@

Quantile regression allows the production of density forecasts without assuming a fixed distributions shape controlled by a small number of parameters, such as the Poisson distribution with parameter $\lambda$. By producing forecasts of multiple quantiles the full predictive distribution can be constructed. Quantile regression is particularly useful where data do not follow a simple distribution, of the distribution shape changes over time of with some covariate, which may be the case with ED arrivals.

Gradient Boosting Machines (GBM) are a tree-based machine learning model for regression and classification. Here we produce probabilistic forecasts of ED arrivals via multiple quantile regression using GBMs as implemented in the R package `gbm` [@Greenwell2020]. GBMs are a best-in-class algorithm for similar regression problems characterised by modest volumes of training data and possible interactions between input features, and are therefore an appealing choice for ED arrival forecasting. Here GBMs are fit with the following features: hour of the day, day of the week, school holiday, temperature and day of the year. Hyperparameters are chosen via grid search on the training data, and were chosen as follows: 500 trees, an interaction depth of 2, and a shrinkage 0.1. An advantage of GBMs is their ability to learn interactions in comparison to the additive models presented later which require the user to specify possible interactions between inputs. Here GBM is included as a reference for the performance of `out-of-the-box' machine learning models applied with minimal effort. We also explored the possibility of performing linear regression with additive models, but these did not perform as well as GBM so are omited for brevity.

@@

Suppose we assume that our predictive distribution follows a given parametric distribution, as in Poisson regression discussed above. In that case, the forecasting task becomes one of predicting the future values of that distribution's parameters. We then can use Generalised Additive Models for Location, Scale and Shape (GAMLSS). These are the distributional regression models where the parameters are modelled as additive functions of explanatory variables. This provides a powerful and flexible framework for probabilistic forecasting, provided that suitable distribution and additive model structures can be found. In practice, this means employing expert judgement and experimenting with various distributions and evaluating their suitability using available training data.

@@

Furthermore, we consider an extension to the additive models for $\lambda_t$ and $\mu_t$ above to incorporate school and public holidays into $D$. These models are labelled Poisson-2 and NOtr-2. We also performed experiments with the truncated $t$ distribution (Ttr-2) and negative binomial distribution (NBI-2), but these did not results in forecasts as well calibrated as the truncated normal.

@@

<!-- <!-- \begin{equation} --> 
<!-- <!--   F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)} --> 
<!-- <!--  (\#eq:truncatedn) --> 
<!-- <!-- \end{equation} -->

@@

<!-- \begin{align*} -->
<!--   \mu_t = &  \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad, \\ -->
<!--   log(\sigma_t) & = f(H(t)) . -->
<!-- \end{align*} -->

@@

To compare model performance, and the significance of any apparent difference in performance, we will use skill scores, which can be calculated for any metric via:
\begin{equation}
  \mathrm{Skill} = \frac{M_\mathrm{ref} - M}{M_\mathrm{ref}} (\#eq:skillscore)
\end{equation}
where $M$ is the metric's value for the method being considered, $M_\mathrm{ref}$ is the metric's value for a reference method. The skill score shows by how many per cent the reference approach is worse than the one under consideration. We will use bootstrap re-sampling of skill scores to determine if the differences in forecast performance (i.e. positive or negative skill) are significantly different from zero [@Efron1981Bootstrap]. Here we use the best performing simple benchmark, Climatology (explained in Subsection \@ref(climatology)), as the reference model and employ a block-bootstrap with blocks of length 24h to account for temporal correlations of the underlying data [@hongyi1996bootstrapping;@Bergmeir2016303].

@@

Figure \@ref(fig:Pinball) presents pinball score aggregated across forecasting horizons for each quantile. It shows that the difference in performance among the models mainly comes from the middle of the distribution and somewhat from the upper tail. There is very little difference in performance for the lower tail (except for TBATS, which has a consistently higher pinball value than the other models). This is interesting and reassuring that the better models are better at probabilities that matter more to decision-makers.

Probabilistic forecasts are evaluated following the principle of *sharpness subject to calibration*, meaning that the sharper forecast is prefered provided that it is calibrated. Mis-calibrated forecasts are unsuitable for use in decision-making, so they should be excluded. Calibration is evaluated visually in Figure \@ref(fig:quantile-bias), which highlights a systematic negative bias across all probability levels in many models, with only the truncated normal and $t$ family GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and ADAM-iETSX models showing good calibration across most probability levels. Notably, both benchmarks exhibit negative quantile bias as they struggle to capture the long term trend of increasing attendance. This could result in poor staffing decisions for a user as historic data fail to characterise present conditions accurately.

@@

<!-- ```{r lead-time-rel, fig.cap= "Pinball values over different forecast horizons.", fig.align='center',out.width= "90%"} -->
<!-- ``` -->

<!-- ```{r Reliability, fig.cap= "Reliability score, showing the relation between empirical and theoretical quantiles.", fig.align='center',out.width= "90%"} -->
<!-- ``` -->

@@

<!--
```{r Skill-rel2bench-reduced, fig.cap= "Pinball skill score for the seven best performing approaches.", fig.align='center',out.width= "70%"}
```

 Figure \@ref(fig:Skill-rel2bench) illustrates the trade-off between calibration and pinball skill. <!-- We may need to expand this-->

<!-- ```{r rmse, fig.cap= "RMSE ...", fig.align='center',out.width= "70%"} -->
<!-- ``` -->

@@

Figure \@ref(fig:lead-time-rmse) reports the RMSE for each forecast horizon. It illustrates the times of day that are harder to predict – morning pick-up and afternoon peak. We can see that some models perform much better than others on specific forecast horizons. For example, Benchmark does an excellent job in predicting 1 - 5 steps ahead ED arrivals and, in general, doing well in forecasting arrivals in the night. In fact, the Benchmark model performs consistently well in terms of the conditional mean, not making as huge mistakes as, for example, Prophet, TBATS and GBM do for some lead times.

<!-- ```{r time, fig.cap= "Running time ...", fig.align='center',out.width= "70%"} -->
<!-- ``` -->

@@

@@

Two approaches produced highly accurate, calibrated probabilistic forecasts: a time series model and a model based on distributional regression. The first is ADAM-iESTX, an extension of exponential smoothing incorporating multiple seasonalities, explanatory variables and assuming a Gamma predictive distribution. The second, labeled NOtr-2, regressed the two parameters of a truncated (at zero) normal distribution on the date and time features and temperature. Both approaches produced calibrated probabilistic forecasts, but the point prediction produced by ADAM-iESTX had a lower RMSE than NOtr-2, while NOtr-2 produced forecasts with a lower pinball score. This suggests that the latter may be preferred if the whole distribution is used in decision-making.

@@

Finally, we have found that out-of-the-box models, which require minimal tuning or manual development, do not perform as well as well-considered statistical approaches. The popular TBATS, Prophet and Gradient Boosting Machine algorithms perform poorly compared to ADAM-iETSX and NOtr-2, and even the benchmarks. Of the models requiring a modest amount of user input and expertise, exponential smoothing (ETS) was found to perform well. ETS produces reasonably well-calibrated forecasts, in contrast to the poorly calibrated benchmarks and has highly accurate point forecasts. However, its probabilistic forecasts were considerably worse than  NOtr-2 in terms of pinball score.

@@

Further research is required to investigate the practical benefits of probabilistic forecasts in healthcare and how they can inform planning and decision making. This may require employing discrete simulation or application to the newsvendor problems. While this study has focused on hourly short-term forecasting, producing longer-term daily forecast (e.g. 180-270 days ahead) is often required by planners to support winter planning in ED and Ambulance services which requires more investigation. Moreover, more research is needed in the forecasting of other important variables such as length of stay, bed occupancy and waiting time, in addition to patient arrivals and admissions. This may require considering the dynamics among various services, including General Practitioners, Emergency Departments, Ambulance and Fire & Rescue services.

>>>>

---
title:  Probabilistic forecasting of hourly Emergency Department arrivals
# author:
#   - name: Bahmna Rostami Tabar
#     email: rostami-tabarb@cardiff.ac.uk
#     affiliation: Cardiff Business School
#     footnote: 12
#   - name: Jethro Browell
#     email: jethro.browell@glasgow.ac.uk
#     affiliation: University of Glasgow
#     footnote: 1
#   - name: Ivan Svetunkov
#     email: i.svetunkov@lancaster.ac.uk
#     affiliation: Lancaster University
#     footnote: 1
# address:
#   - code: Cardiff Business School
#     address: Cardiff Business School, Cardiff University, UK
#   - code: University of Glasgow
#     address: School of Mathematics and Statistics, University of Glasgow, UK
#   - code: Lancaster University
#     address: Centre for Marketing Analytics and Forecasting, Lancaster University, UK
# footnote:
#   - code: 12
#     text: "Corresponding Author"
#   - code: 1
#     text: "Equal contribution"
abstract: |
   An accurate forecast of Emergency Department (ED) arrivals by an hour of the day is critical to meet patients’ demand. It enables planners to match ED staff to the number of arrivals, redeploy staff, and reconfigure units. This can have many advantages for healthcare staff and the quality of care delivered to patients. In this study, we develop an innovative model based on Generalised Additive Models and an advanced dynamic model based on exponential smoothing to generate an hourly probabilistic forecast of ED arrivals for a prediction window of 48 hours. We compare the forecast accuracy of these models against appropriate benchmarks, including TBATS, Poisson Regression, Prophet, and simple empirical distribution. We use Root Mean Squared Error (RMSE) to examine the point forecast accuracy and assess the forecast distribution accuracy using Quantile Bias, PinBall Score and Pinball Skill Score. Our results indicate that the proposed models outperform their benchmarks for point and probabilistic forecasts. Our developed models can also be generalised to forecast hourly arrivals in other services such as hospitals, ambulances, or clinical desk services.
keywords:  "Emergency Department, Poisson Regression, Probabilistic Forecasting, Generalised Additive Models, Intermittent Exponential Smoothing"

journal: "Annals of Emergency Medicine"
date: "`r Sys.Date()`"
geometry: "top=25mm, left=25mm, right=25mm, bottom=25mm,headsep=10mm, footskip=12mm"
#linenumbers: true
numbersections: true
#output: rticles::elsevier_article
header-includes:  
 \usepackage{adjustbox, float,lscape,amsmath,bm} #use the 'float' package
output:
  bookdown::pdf_book: 
    base_format: rticles::elsevier_article
    citation_package: natbib
bibliography: mybibfile.bib
#biblio-style: authoryear
#csl: elsevier-without-titles.csl
#biblio-style: abbrevnat
editor_options: 
  markdown: 
    wrap: 72
---

@@

There are few studies that look at forecasting hourly arrivals in ED and
other hospital services using historical time series data and/or
predictors such as patient characteristics, weather, holidays and public
events. These studies use multiple approaches, including Exponential
Smoothing [@SvetunkovAdam], Autoregressive Integrated Moving Average
(ARIMA) [@hyndman2021forecasting], Autoregressive Conditional
Heteroskedasticity (ARCH) [@bollerslev1994arch], Vector Autoregressive
model [@lutkepohl2013vector], TBATS [@de2011forecasting] and Artificial
Neural Networks [@hyndman2021forecasting]. \textcolor{blue}{We have identified a number of limitations and gaps in the area of ED forecasting that motivated us to undertake this study. These gaps and their importance are discussed below.} \textcolor{blue}{Most of these studies are limited to only predicting future arrivals as a point
forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.} There are few studies that report uncertainty by presenting prediction intervals, but \textcolor{blue}{there is no study generating and evaluating the entire forecast distribution of arrivals. Reporting the uncertainty via the forecast distribution is potentially valuable in this setting and has practical implications for those managing Emergency Departments because the consequences of inadequate staffing are asymmetric}. This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient. Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently. Furthermore, if the impact of under- and overstaffing can be quantified,
probabilistic forecasts allow _optimal_ decisions that balance the cost associated with under- and overstaffing to be calculated. Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task. \textcolor{blue}{Another drawback of existing studies is that the
datasets used are relatively small (e.g. time period of 1-2 years), making it challenging to report the forecast accuracy using robust approaches such as time series cross-validation. Such results might not be generalisable. Additionally, most of the forecasting methods used in these publications do not consider the full extent of the multiple seasonality of hourly ED arrivals. Moreover, hourly ED time series may contain low volume values and potentially zeros in some hours of the day, which brings some challenges to traditional time series forecasting resulting in producing negative forecasts of future arrivals.
Finally, all previous publications
referenced in this paper are not fully reproducible as underlying data and code are not available.}

@@

# Research background: hourly ED forecasting {#lit}
\textcolor{blue}{
There is a substantial number of studies that employ models to forecast admissions and arrivals to inform planning and decision making in the healthcare. Areas such as forecasting call volume arrivals, ambulance demand and Emergency Department forecasting has received a significant attention. We refer interested readers to  some extensive reviews of the relevant literature} [@mingliterature2022;@gul2020review;@ibrahim2016modeling]. \textcolor{blue}{The time granularity considered by these studies spans from hourly to yearly across different parts of the healthcare. However, given
the focus of this study, we only discuss hourly ED forecasting.}

@@

Table \@ref(tab:summarylit) summarises the relevant papers.

@@

According to the studies mentioned above, it can be said that they have
shown complications in forecasting hourly patient accident and emergency
department visits, and the application of forecasting hourly patient
visits is not well established. Some of the studies claimed that the
accuracy of forecasting models on hourly accident and emergency
department data is low [@boyle2012predicting; @hertzum2017forecasting],
while others mentioned that the accuracy of ED hourly forecast is at an
acceptable level [@choudhury2020forecasting; @mccarthy2008challenge;
@schweigler2009forecasting].

@@

Data used in this study comprises counts of patients' arrival times at
one of the largest ED units in the UK between April 2014 and February
2019, extracted from the ED administrative database of the hospital. We
aggregated the patients' arrival times to obtain hourly arrivals, which
are used in this study. Figure \@ref(fig:hourly-plot-ridge) illustrates
the distribution of arrivals for each hour of the day and the day of the
week. Although the data is noisy, it reveals some systematic patterns.

@@

```{r seasonplot-dofw, fig.cap= "Subseries plot: day of weekly arrivals", fig.align='center',out.width="70%"}
```

Figure \@ref(fig:seasonplot-dofw) \textcolor{blue}{illustrates the daily subseries plot, with the x-axis representing the date and y-axis the ED arrivals. Each individual plot illustrates how arrivals change over time for a each day of week (i.e. seven days) from Monday to Saturday. The blue line shows the average arrival for the given day. It is clear that ED arrivals on Monday are higher followed by Saturday. This indicates that there are
significantly more arrivals on Mondays and Saturdays compared to the rest of the week. This might be due to the closure of General Practitioners outpatient clinics over the weekend.}

@@

## Naive {#naive}

We start with one of the simplest forecasting approaches used in
practice - assuming that in the next few hours, everything will be the
same as in the similar hours of a similar day in the past. This is
called "Naïve". In our case, given that we need a distribution of
values, we will use a modified approach, where the
empirical distribution of the hourly arrival time series is used to
forecast the future arrival distribution [@la2021new]. We consider the
empirical distribution of all available historic data (Benchmark-1) and
the empirical distribution of the most recent year of historic data on a
rolling basis (Benchmark-2) to capture potential changes in behaviour
over time.

@@

The variables _1--3_ allow modelling the seasonal patterns on the
appropriate level of detail throughout the year, while _4_ covers the
changes in admittance due to calendar events. Finally, _5_ is needed to
capture the potential phenomenon of change in admittance after the
holiday (e.g. people might try not to go to the hospital on Christmas
eve and thus go the next day). This model assumes that all these effects
are deterministic and do not change over time. Still, the exponentiation
in \@ref(eq:PoissonRegression) introduces an interaction effect between
dummy variables so that the 3 pm on Monday in January will be different
from 3 pm on Monday in July, although the parameters for the hour of day
and day of the week are fixed and do not change over time. We use the
`alm()` function from the `greybox` package [@Svetunkov2021Greybox] for
R [@RTeam2021] for the experiments and denote this model as "Poisson
Regression".

@@

Prophet is a forecasting procedure created by Facebook
[@taylor2018forecasting] that accounts for multiple seasonality,
piecewise trend and holiday effects. Prophet is robust to missing data
and shifts in the trend and typically handles outliers well. Prophet
works well on daily data seen in Facebook. It is robust and automated,
making it easy to learn for beginners. The implementation may be less
flexible than other methods. The model itself relies on the Multiple
Source of Error state space model, initially proposed by @Kalman1960.
The model is incorporated using the corresponding implementation of the
Fable package in R. We use the `prophet()` function from the `fable`
package [@fable2020]. Note that the input data is assigned with an
hourly and daily seasonality. \textcolor{blue}{This method has been adopted in some healthcare service providers in the United Kingdom to produce forecasts, therefore we have included it as one of our benchmarks.}

@@

Quantile regression allows the production of density forecasts without
assuming a fixed distributions shape controlled by a small number of
parameters, such as the Poisson distribution with parameter $\lambda$.
By producing forecasts of multiple quantiles the full predictive
distribution can be constructed. Quantile regression is particularly
useful where data do not follow a simple distribution, the
distribution shape changes over time with some covariate, which may
be the case with ED arrivals.

Gradient Boosting Machines (GBM) are a tree-based machine learning model
for regression and classification. Here we produce probabilistic
forecasts of ED arrivals via multiple quantile regression using GBMs as
implemented in the R package `gbm` [@Greenwell2020]. GBMs are a
best-in-class algorithm for similar regression problems characterised by
modest volumes of training data and possible interactions between input
features, and are therefore an appealing choice for ED arrival
forecasting. Here GBMs are fit with the following features: hour of the
day, day of the week, school holiday, temperature and day of the year.
Hyperparameters are chosen via grid search on the training data, and
were chosen as follows: 500 trees, an interaction depth of 2, and a
shrinkage 0.1. An advantage of GBMs is their ability to learn
interactions in comparison to the additive models presented later which
require the user to specify possible interactions between inputs. Here
GBM is included as a reference for the performance of \`out-of-the-box'
machine learning models applied with minimal effort. We also explored
the possibility of performing linear regression with additive models,
but these did not perform as well as GBM so are omited for brevity.

@@

Suppose we assume that our predictive distribution follows a given
parametric distribution, as in Poisson regression discussed above. In
that case, the forecasting task becomes one of predicting the future
values of that distribution's parameters. We then can use Generalised
Additive Models for Location, Scale and Shape (GAMLSS). These are the
distributional regression models where the parameters are modelled as
additive function of explanatory variables. This provides a powerful
and flexible framework for probabilistic forecasting, provided that
suitable distribution and additive model structures can be found. In
practice, this means employing expert judgement and experimenting with
various distributions and evaluating their suitability using available
training data.

@@

Furthermore, we consider an extension to the additive models for
$\lambda_t$ and $\mu_t$ above to incorporate school and public holidays
into $D$. These models are labelled Poisson-2 and NOtr-2. We also
performed experiments with the truncated $t$ distribution (Ttr-2) and
negative binomial distribution (NBI-2), but these did not result in
forecasts as well calibrated as the truncated normal.

@@

<!-- <!-- \begin{equation} -->

<!-- <!--   F_t(y_t,\mu_t,\sigma_t) =  \frac{\Phi\left( \frac{y_t-\mu_t}{\sigma_t} \right) - \Phi\left( \frac{-y_t}{\sigma_t} \right)}{1 - \Phi\left( \frac{-y_t}{\sigma_t} \right)} -->

<!-- <!--  (\#eq:truncatedn) -->

<!-- <!-- \end{equation} -->

@@

<!-- \begin{align*} -->

<!--   \mu_t = &  \sum_{i=1}^7 \beta_i \delta(D_i(t)-i) + \sum_{j=1}^7 D_j(t) f_j(H(t)) + t f_\text{Y}(Y(t)) + f_\text{Temp}(Y(t),C_t) \quad, \\ -->

<!--   log(\sigma_t) & = f(H(t)) . -->

<!-- \end{align*} -->

@@

To compare model performance, and the significance of any apparent
difference in performance, we will use skill scores, which can be
calculated for any metric via: \begin{equation}
  \mathrm{Skill} = \frac{M_\mathrm{ref} - M}{M_\mathrm{ref}} (\#eq:skillscore)
\end{equation} where $M$ is the metric's value for the method being
considered, $M_\mathrm{ref}$ is the metric's value for a reference
method. The skill score shows by how many per cent the reference
approach is worse than the one under consideration. We will use
bootstrap re-sampling of skill scores to determine if the differences in
forecast performance (i.e. positive or negative skill) are significantly
different from zero [@Efron1981Bootstrap]. Here we use the best
performing simple benchmark, Naive (explained in Subsection
\@ref(naive)), as the reference model and employ a block-bootstrap
with blocks of length 24h to account for temporal correlations of the
underlying data [@hongyi1996bootstrapping; @Bergmeir2016303].

@@

Figure \@ref(fig:Pinball) presents Pinball score aggregated across
forecasting horizons for each quantile. It shows that the difference in
performance among the models mainly comes from the middle of the
distribution and somewhat from the upper tail. There is very little
difference in performance for the lower tail (except for TBATS, which
has a consistently higher Pinball value than the other models). This is
interesting and reassuring that the better models are better at
probabilities that matter more to decision-makers.

Probabilistic forecasts are evaluated following the principle of
*sharpness subject to calibration*, meaning that the sharper forecast is
prefered provided that it is calibrated. Mis-calibrated forecasts are
unsuitable for use in decision-making, so they should be excluded.
Calibration is evaluated visually in Figure \@ref(fig:quantile-bias),
which highlights a systematic negative bias across all probability
levels in many models, with only the truncated normal and $t$ family
GAMLSS models (NOtr-1, NOtr-2, Ttr-2) and ADAM-iETSX models showing good
calibration across most probability levels. \textcolor{blue}{ Notably, both benchmarks
exhibit negative quantile bias as they struggle to capture the long term trend of increasing arrivals. This could result in poor staffing decisions. This is because the empirical distribution of whole data fails to characterise how arrivals in ED may change over time.}

@@

<!-- ```{r lead-time-rel, fig.cap= "Pinball values over different forecast horizons.", fig.align='center',out.width= "90%"} -->

<!-- ``` -->

<!-- ```{r Reliability, fig.cap= "Reliability score, showing the relation between empirical and theoretical quantiles.", fig.align='center',out.width= "90%"} -->

<!-- ``` -->

@@

````{=html}
<!--
```{r Skill-rel2bench-reduced, fig.cap= "Pinball skill score for the seven best performing approaches.", fig.align='center',out.width= "70%"}
```

 Figure \@ref(fig:Skill-rel2bench) illustrates the trade-off between calibration and Pinball skill. <!-- We may need to expand this-->
````

<!-- ```{r rmse, fig.cap= "RMSE ...", fig.align='center',out.width= "70%"} -->

<!-- ``` -->

@@

Figure \@ref(fig:lead-time-rmse) reports the RMSE for each forecast
horizon. It illustrates the times of day that are harder to predict --
morning pick-up and afternoon peak. We can see that some models perform
much better than others on specific forecast horizons. For example,
Benchmark does an excellent job in predicting 1 - 5 steps ahead ED
arrivals and, in general, doing well in forecasting arrivals in the
night. In fact, the Benchmark model performs consistently well in terms
of the conditional mean, not making as huge mistakes as, for example,
Prophet, TBATS and GBM do for some lead times.

<!-- ```{r time, fig.cap= "Running time ...", fig.align='center',out.width= "70%"} -->

<!-- ``` -->

@@

\textcolor{blue}{The benefits of the proposed model and its managerial aspects include
the use of probabilistic forecast to inform decision making and provide
insights on the running time of each model to generate forecasts. Probabilistic forecasts contain all potential future outcomes and help
planners to achieve a more efficient decision making by not only allocate a probability to the most likely outcome but also to unlikely
outcomes such as extremely high or low arrivals in the Emergency Departments. The running time could be also an important aspects for managers depending on the frequency of generating forecasts. If the frequency of forecast generation is high, one may employ models such as
Poisson regression (Poisson 1). If the frequency is lower and running time is not a big concern, a model like GAMLSS (Ttr-2) should be used.}

\textcolor{blue}{Probabilistic forecasts can be considered as a useful tool for planners to manage the risk of under and over resource allocation, which consequently may reduce both costs and risks of service delivery in Emergency Departments.}

@@

Two approaches produced highly accurate, calibrated probabilistic
forecasts: a time series model and a model based on distributional
regression. The first is ADAM-iESTX, an extension of exponential
smoothing incorporating multiple seasonalities, explanatory variables
and assuming a Gamma predictive distribution. The second, labeled
NOtr-2, regressed the two parameters of a truncated (at zero) normal
distribution on the date and time features and temperature. Both
approaches produced calibrated probabilistic forecasts, but the point
prediction produced by ADAM-iESTX had a lower RMSE than NOtr-2, while
NOtr-2 produced forecasts with a lower Pinball score. This suggests that
the latter may be preferred if the whole distribution is used in
decision-making.

@@

Finally, we have found that out-of-the-box models, which require minimal
tuning or manual development, do not perform as well as well-considered
statistical approaches. The popular TBATS, Prophet and Gradient Boosting
Machine algorithms perform poorly compared to ADAM-iETSX and NOtr-2, and
even the benchmarks. Of the models requiring a modest amount of user
input and expertise, exponential smoothing (ETS) was found to perform
well. ETS produces reasonably well-calibrated forecasts, in contrast to
the poorly calibrated benchmarks and has highly accurate point
forecasts. However, its probabilistic forecasts were considerably worse
than NOtr-2 in terms of Pinball score.

\textcolor{blue}{
The dataset used in this study does not include the COVID-19 pandemic. During COVID-19, the dynamic of ED has changed. That means any forecasting model that is used during such periods to forecast ED arrivals needs to be modified to reflect those changes for that specific period. However, we expect that if the COVID period is taken into account by the models (or interpolated), the models should perform similarly.}

@@

Further research is required to investigate the practical benefits of
probabilistic forecasts in healthcare and how they can inform planning
and decision making. This may require employing discrete event simulation or
application to the newsvendor problems. While this study has focused on
hourly short-term forecasting, producing longer-term daily forecast
(e.g. 180-270 days ahead) is often required by planners to support
winter planning in ED and Ambulance services which requires more
investigation. Moreover, more research is needed in the forecasting of
other important variables such as length of stay, bed occupancy and
waiting time, in addition to patient arrivals and admissions. This may
require considering the dynamics among various services, including
General Practitioners, Emergency Departments, Ambulance and Fire &
Rescue services.
